Sediment transport capacity response to variations in water discharge in pressurized subglacial channels

### 2025-02-07 Delaneyetal2024.tex
small changes

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In full subaerial channels, water discharge variations are accommodated by flow depth and width changes, along with water velocity.
  However, in subglacial channels, water is pressurized by the ice above, and they grow in response to frictional heating of water flowing through them. 
  As a result, rapid changes in water discharge mainly result in velocity variations, as the channel geometry evolves slowly.
  Here, we present formulations of sediment transport capacity in different channel types and apply subglacial and subaerial hydraulics models to hydrographs from an Alpine glacier and the Greenland Ice Sheet. 
  Numerical experiments show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  This hysteresis causes a highly variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  Reducing water discharge variability by smoothing lessens the hysteresis effects, in some cases to the point where a covarying relationship between water discharge and sediment transport capacity may occur, similar to subaerial systems.
  A second set of numerical experiments shows that subglacial sediment transport is highly non-linear with respect to water discharge, creating more variability in sediment transport capacity.
  Yet, results and formulations of subglacial sediment transport capacity show that its variability can approach that of subaerial systems when subglacial channel size is in equilibrium with water discharge.
  We present these results in the context of different formulations of the relationship between water discharge and sediment transport capacity-based assumptions of subglacial and subaerial flow regimes.
  The implications of these findings help to evaluate sediment discharge from glaciers with different hydro-climatic forcings.
  These findings can improve the interpretation of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Despite the different physical processes between the subaerial and subglacial systems, observed water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, especially given the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited regime, instead of sediment availability.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchments \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019b} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

@@

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W), at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012 \citet[][Figure~\ref{fig:model_outs}\,b]{tedstone2013}.

@@

Different hydraulic gradients control the velocity and sediment transport capacity in the subglacial and subaerial cases (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}). 
The subglacial one is generally much steeper \citep{alley1997}.
This results in the shear stresses and width-integrated shear stresses across the channel bed that controls sediment transport capacity being much greater in subglacial channels (Figure~\ref{fig:model_outs}).
The parameters' range tested in Section~\ref{sect:ensemble} covers a likely span of viable shear stresses in both subglacial and subaerial channels.
The greater sediment transport capacity in the subglacial channels implies that sediment grain size underneath subglacial channels is likely larger than subaerial counterparts.
Note, however, that the water velocities are similar in both channel types.

@@

\FloatBarrier
\subsection{Sediment transport scaling in different channel types}
\label{sect:scaling}

@@

The numerical experiments above consider the size evolution of subglacial channels and demonstrate that for these hydrographs subglacial sediment transport variability is greater than its subaerial counterpart (Section~\ref{sect:ensemble}). 
Additionally, results demonstrate the impact of water discharge variability on sediment transport capacity.
Here, we discuss and compare the sediment transport behavior of different channel types as they respond to water discharge, channel shape, and hydraulic gradient, extending the formulations presented in \citet{alley1997}.

@@

Catchments with reduced water discharge variability may experience less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and variable relationship between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge (Figure~\ref{fig:chan_types} b).
High water discharge variability in \alpine{} may cause width integrated shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Figure~\ref{fig:chan_types}; Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} comes as a result of the larger catchment areas and longer travel times of the water \citep[e.g.][]{vanas2017}.
Water discharge's stronger correlation with width-integrated shear stress in \icesheet{} may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge (Figure~\ref{fig:chan_types} d).
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The manuscript's first objective is to establish the conditions where water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak water velocity and sediment transport capacity occurs before peak water discharge during a discharge event on hydrographs from an alpine and ice sheet catchment, due to evolving channel size.
In subaerial channels, the timing of peak sediment transport capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, a variable relationship between water and subglacial sediment discharge.
Water discharge variations in both Greenland Ice Sheet and Alpine are substantial enough to cause this inconsistent relationship, even when water discharge records are smoothed.
This variable relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Water discharge and sediment transport capacity covariance could be possible when water discharge varies at a slower rate than subglacial channel size, such as Antarctica with minimal surface melt input.
Results also suggest that water discharge records averaged over periods longer than $12$ hours from alpine and ice sheet hydrographs show substantial impacts on subglacial sediment transport capacity characteristics.

@@

Few observations of the subglacial environment hinder our ability to quantify processes such as
the shape of subglacial channels and the response time of subglacial channels to water discharge variations.
The poor constraints on subglacial sediment size and sorting processes make it difficult to link shear stress to sediment transport capacity.
New methods are being developed to quantify these processes \citep{jenkin2023},  better informing the response of sediment transport capacity to water discharge forcing in subglacial environments.

>>>>

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In full subaerial channels, water discharge variations are accommodated by flow depth and width changes, along with water velocity.
  However, in subglacial channels, water is pressurized by the ice above, and they grow in response to frictional heating of water flowing through them. 
  As a result, rapid changes in water discharge mainly result in velocity variations, as the channel geometry evolves slowly.
  Here, we present formulations of sediment transport capacity in different channel types and apply subglacial and subaerial hydraulics models to hydrographs from an Alpine glacier and a catchment of the Greenland Ice Sheet. 
  Numerical experiments show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  This hysteresis causes a highly variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  Reducing water discharge variability by smoothing lessens the hysteresis effects, in some cases to the point where a covarying relationship between water discharge and sediment transport capacity may occur, similar to subaerial systems.
  A second set of numerical experiments shows that subglacial sediment transport is highly non-linear with respect to water discharge, creating more variability in sediment transport capacity.
  Yet, results and formulations of subglacial sediment transport capacity show that its variability can approach that of subaerial systems when subglacial channel size is in equilibrium with water discharge.
  We present these results in the context of different formulations of the relationship between water discharge and sediment transport capacity-based assumptions of subglacial and subaerial flow regimes.
  The implications of these findings help to evaluate sediment discharge from glaciers with different hydro-climatic forcings.
  These findings can improve the interpretation of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Despite the different physical processes between the subaerial and subglacial systems, observed water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, especially given the role of sediment availability or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited regime.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchments \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019b} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

@@

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W), at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012 \citep[][; Figure~\ref{fig:model_outs}\,b]{tedstone2013}.

@@

Different hydraulic gradients control the velocity and sediment transport capacity in the subglacial and subaerial cases (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}). 
The subglacial one is generally much steeper \citep{alley1997}.
This results in the shear stresses and width-integrated shear stresses across the channel bed that controls sediment transport capacity being much greater in subglacial channels (Figure~\ref{fig:model_outs}).
The parameters' range tested in Section~\ref{sect:ensemble} covers a likely span of viable shear stresses in both subglacial and subaerial channels.
The greater sediment transport capacity in the subglacial channels implies that sediment grain size underneath subglacial channels is likely larger than subaerial counterparts, as smaller clasts would be evacuated.
Note, however, that the mean water velocities are similar in both channel types.

@@

\FloatBarrier
\subsection{Sediment transport scaling in different channel types}
\label{sect:scaling}
The numerical experiments above consider the size evolution of subglacial channels and demonstrate that for these hydrographs subglacial sediment transport variability is greater than its subaerial counterpart (Section~\ref{sect:ensemble}). 
Additionally, results demonstrate the impact of water discharge variability on sediment transport capacity.
Here, we discuss and compare the sediment transport behavior of different channel types as they respond to water discharge, channel shape, and hydraulic gradient, extending the formulations presented in \citet{alley1997}.

@@

@@

Catchments with reduced water discharge variability may experience less variability in sediment transport capacity, shown by the \alpine{} and \icesheet{} model outputs.
Indeed, the more decoupled and variable relationship exists between model outputs and water discharge in \alpine{} results from the relatively larger discharge variations on sub-daily to weekly timescales (Figure~\ref{fig:Qw_vari}).
This occurs as the subglacial channel's size evolves slowly compared to the variations in water discharge (Figure~\ref{fig:chan_types} b).
High water discharge variability in \alpine{} may cause width integrated shear stress to approach $Q^{2}$ in assuming pipe-flow conditions where the water discharge varies largely compared to channel size \citep[Figure~\ref{fig:chan_types}; Section \ref{sect:scaling}; c.f.][]{alley1997}.
Conversely, the reduced relative variability in water discharge in \icesheet{} comes as a result of the larger catchment areas and longer travel times of the water \citep[e.g.][]{vanas2017}.
Water discharge's stronger correlation with width-integrated shear stress in \icesheet{} may result in the subglacial channel's size being closer to equilibrium from the smaller variations in water discharge (Figure~\ref{fig:chan_types} d).
In this case, the exponent on water discharge for shear stress is likely substantially less than $w\tau \propto Q^2$, but greater than $w \tau \propto Q^{\frac{4}{5}}$ that occurs in a steady-state R-channel (see Section \ref{sect:scaling}).
As a result, there is a stronger relationship between subglacial model outputs and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}) and less variability model outputs in the \icesheet case (Figure~\ref{fig:multi_run}).

@@

The manuscript's first objective is to establish the conditions where water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak water velocity and sediment transport capacity occurs before peak water discharge during a discharge event on hydrographs from an alpine and ice sheet catchment, due to evolving channel size.
In subaerial channels, the timing of peak sediment transport capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, a variable relationship between water and subglacial sediment discharge.
Water discharge variations in potentially  representative Greenland Ice Sheet and Alpine catchments are substantial enough to cause this inconsistent relationship, even when water discharge records are smoothed.
This variable relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Water discharge and sediment transport capacity covariance could be possible when water discharge varies at a slower rate than subglacial channel size, such as Antarctica with minimal surface melt input.
Results also suggest that water discharge records averaged over periods longer than $12$ hours from alpine and ice sheet hydrographs show substantial impacts on subglacial sediment transport capacity characteristics.

@@

Few observations of the subglacial environment hinders our ability to quantify processes such as
the shape of subglacial channels and the response time of subglacial channels to water discharge variations.
The poor constraints on subglacial sediment size and sorting processes make it difficult to link shear stress to sediment transport capacity.
New methods are being developed to quantify these processes \citep{jenkin2023},  better informing the response of sediment transport capacity to water discharge forcing in subglacial environments.
